Modelling with ANIMO:\\ between Fuzzy Logic and Differential Equations

### 2015-06-29 Paper.tex
Corrected minor errors, updated Figure 5 with labels for hypotheses, written cover letter

\subsection*{A brief introduction to ANIMO}
The user interface of ANIMO is displayed in Figure~\ref{fig:cytoscape},
where we present an example of a biological network enriched with dynamic information.
ANIMO is implemented as a plug-in to Cytoscape~\cite{cytoscape}, a software tool developed to represent biological networks.
On top of the static topological information displayed in Cytoscape, ANIMO represents
biological interactions starting from the basic concept of \emph{activity}:
each biological entity in an ANIMO model is considered to be either active or inactive.
Activity is to be interpreted in a very broad sense: for example, an active gene is being transcribed,
an active kinase can perform phosphorylations, etc. Each node in an ANIMO network represents
both active and inactive entities of the same type, with the relative amount of active entities
(the \emph{activity level}) represented by the node colour on a user-configurable scale.
Interactions among nodes define how the biological entities in a network influence each other's activity.
Only nodes whose activity level is larger than 0 (such nodes are called \emph{active}) can have an 
influence on their downstream targets, and only if that influence is not counterbalanced by intervening opposite interactions.
For example, the interaction $A\rightarrow B$ (read ``A activates B'') indicates that node A, if active,
will increase the activity of node B. If we add an additional interaction to the example,
$C \dashv$\hspace{0.1em}$B$ (``C inhibits B''), with C also active, then the activity level of B will change depending on the
activity levels of A and C, and on their quantitative influence.
The influence of an ANIMO interaction is quantified by a scale factor $k$, which defines the speed at which
that interaction occurs: higher values of $k$ give faster interactions. These $k$-values are
the only parameters needed in an ANIMO model, and can be given in either
a quantitative (as real numbers) or a qualitative way, choosing among self-explanatory descriptions
such as ``very slow'', ``slow'', ``medium'', ``fast'', ``very fast''.

@@

\section*{Results}\label{sec:results}
\subsection*{Modelling oscillation in Drosophila Melanogaster circadian clock}\label{sec:animo-drosophila}
To demonstrate that results obtained with ANIMO are comparable to results from widely used modelling
approaches, we present 
an ANIMO model of the circadian clock in \emph{Drosophila Melanogaster} (Fig.~\ref{fig:drosophila-model}).
This ANIMO model is based on~\cite{drosophila-ode-model},
where ordinary differential equations (ODEs) were used.
The cyclic behaviour of the circadian clock is based on the alternating formation and destruction of the
CYC/CLK protein complex.
Concentration levels of this complex are in turn regulated by a series of proteins which are produced as
a consequence of CYC/CLK formation. The CWO protein
is central to the functioning of the network, as it degrades the mRNA for most of the involved proteins.
As such, CWO acts as an inhibitor that counterbalances the effect of CYC/CLK.
The positive influence of the light-regulated cryptochrome CRY on the degradation of TIM is a consequence
of the passage between day and night, allowing
the circadian clock to synchronize to a time zone (see Suppl. Sect.~\ref{suppl:repressilator}).

@@

\begin{figure*}[!htpb]
\begin{center}
% \subfloat[\label{fig:large-model-complete}]{\includegraphics[scale=\largeModelScale]{images/large_network_legendg}}\\
%  \subfloat[\label{fig:large-model-graph2}]{\includegraphics[scale=\scalaGrafici]{images/TGF100_vs_TNF100_hyp1_IL-1a4}}
%  \subfloat[\label{fig:large-model-graph4}]{\includegraphics[scale=\scalaGrafici]{images/TNF5_C225_hyp2_JNK1_MK2}}
%  \includegraphics[width=0.9\textwidth]{images/large_network_legendg_completa}
  \includegraphics[width=0.9\textwidth]{Figures/5}
\end{center}
\caption{\csentence{Signalling network downstream of TNF$\alpha$ and EGF in human colon carcinoma cells: improved model.}
{\bf(A)} %{\bf \protect\subref{fig:large-model-complete}}
The model for the merged TNF$\alpha$ and EGF pathways
after addition of the two hypotheses (highlighted).
Hypothesis 1 assumes IL-1$\alpha$ expression to depend on AP-1 activity, which in turn requires
both c-Jun en c-Fos to be activated by JNK1 and ERK, respectively. Hypothesis 2 assumes RAS as an activator
of MEKK1. Node colours represent the activity levels $15$ minutes
after stimulation of 100 ng/ml TNF$\alpha$ + 100 ng/ml EGF.
{\bf(B)} %{\bf \protect\subref{fig:large-model-graph2}}~
After the addition of the first hypothesis (activation of IL-1$\alpha$ production depending both
on JNK1 and ERK): production of IL-1$\alpha$ after stimulation with 100 ng/ml TNF$\alpha$ (series {\sf IL-1a~(TNFa)})
compared with stimulation with 100 ng/ml TGF$\alpha$ (series {\sf IL-1a~(TGFa)}) (24 hours).
{\bf(C)} %{\bf \protect\subref{fig:large-model-graph4}}~
After the addition of the second hypothesis (activation of MEKK1 downstream of EGFR):
stimulation with 5 ng/ml TNF$\alpha$ + 10 $\mu$g/ml C225 (2 hours).
Suppl. Sect.~\ref{suppl:parameters-tnf-egf} explains how the dosage of 5 ng/ml TNF$\alpha$ was represented in the model.\\
The {\sf \_{}data} suffix identifies experimental data; all other series are computed by ANIMO.}\label{fig:large-model-complete}
\end{figure*}

As a second example, we considered the behaviour of JNK1 and MK2. In the model, both
proteins were located downstream of TNF$\alpha$ but not TGF$\alpha$ or EGF. Hence, the
model did not show an effect of C225, a pharmacological inhibitor of ligand-EGFR
binding, on activation of JNK1 or MK2 after stimulation with TNF$\alpha$. However, experimental
data show that C225 strongly reduces activation of JNK1 and MK2 upon stimulation with
TNF$\alpha$~\cite{pathway-autocrine}.
This fact is indicative of a role for EGFR in activation of JNK1 and MK2. Since both JNK1 and MK2
are located downstream of MEKK1, we hypothesized that activation
of MEKK1 is dependent on
both TNF$\alpha$-signalling and TGF$\alpha$-signalling. In the model we added a new
hypothetical node {\sf Hyp~2} (hypothesis~2) to link EGFR to MEKK1 (Suppl. Fig.~\ref{fig:large-model-hypotheses}).
This addition led to an improved fit of the model to the data upon treatment with TNF$\alpha$ + C225:
activation of both MK2 and JNK1 was strongly suppressed by C225 (Fig.~\ref{fig:large-model-complete}C).
Stimulation with EGF alone did not lead to activation of JNK1 and MK2.
These data support the validity of the modification to the model.
Further support for a link between EGFR and MEKK1 was found in literature. Specifically,
Ras has been reported as a direct activator of
MEKK1~\cite{ras-mekk1}. EGFR is a well-known and potent activator of Ras,
which is why it was already in our network~\cite{kegg}.
Other studies also report activation of JNK1 and phosphorylation of c-Jun downstream of Ras, which is consistent with
an interaction between Ras and MEKK1~\cite{cfos-cjun,ras-jnk1}.
Based on these findings, we adapted
our model by removing the {\sf Hyp~2} node and creating a direct interaction between Ras
and MEKK1 (Fig.~\ref{fig:large-model-complete}A). Experimentally, the role of Ras could be confirmed by using a
pharmacological inhibitor of Ras activity, and measuring the effect of this inhibitor on the activation of JNK1 and MK2.
Together, our model suggests that EGFR activity is required
but not sufficient for activation of JNK1 and MK2 in HT-29 cells.

@@

From the point of view of model precision, we position ANIMO between fuzzy logic and ODEs.
Being less parameter-intensive than ODEs more precise than logic-based models,
ANIMO models are flexible enough to be useful in many occasions.

@@

\def\modelGraphScale{0.148}%0.215}
\def\legendGraphScale{0.16}
\def\halfGraphScale{0.067}%0.1075}
\begin{figure*}[!h]
%\begin{floatingfigure}[l]{.45\textwidth}
\begin{center}
% \subfloat[\label{subfig:mek-erk}]{\includegraphics[scale=.098]{images/abstraction_ta_mek-erk3}}
%   \quad%\qquad\qquad
% \subfloat[\label{subfig:erk}]{\includegraphics[scale=.098]{images/abstraction_ta_erk}}
%   \quad%\qquad\qquad
% \subfloat[\label{subfig:mek}]{\includegraphics[scale=.098]{images/abstraction_ta_mek}}
%   \includegraphics[width=0.7\textwidth]{images/abstraction_ta_mek-erk3_oneImage}
  \includegraphics[width=0.8\textwidth]{Figures/6}
\end{center}
\caption{\csentence{Abstraction of a biochemical reaction to a \tas\ model.}
{\bf(A)} %{\bf \protect\subref{subfig:mek-erk}}~
Classical depiction of a well-studied intracellular signal transduction reaction: protein
MAPK-ERK kinase (MEK) activates downstream protein extracellular-regulated kinase (ERK).
{\bf(B)} %{\bf \protect\subref{subfig:erk}}~
A \ta\ model of ERK, consisting of two states (circles), {\sf inactive\_ERK} and {\sf active\_ERK},
and one transition (arrow) between the states. This transition will take place when it is possible to synchronize with
the corresponding action {\sf activate\_ERK!} in the MEK automaton.
{\bf(C)} %{\bf \protect\subref{subfig:mek}}~
A \ta\ model of active MEK, consisting of one state and one
transition. ${\sf t} < 20$ is called an invariant on the state, allowing residence in this state as long as local
clock time {\sf t} is smaller than $20$ units. ${\sf t} > 18$ is called a guard on the transition, allowing the
transition to take place when local clock {\sf t} is greater than $18$ units. Together, the invariant and guard in this
example ensure that the transition must take place in the (continuous) time interval $18 < {\sf t} < 20$. When the
transition takes place, the action {\sf activate\_Erk!} is performed (thus allowing the ERK automaton to reach the {\sf
active\_ERK} state) and the local clock coupled to this automaton is reset, ${\sf t} := 0$.\label{fig:abstraction-mek-erk}}
\end{figure*}

@@

Timed Automata have been shown to be a powerful formalism to model biological processes
~\cite{ta-siebert,bartocci-oscillators,oded-ode-ta-discretization}. A timed automaton consists of locations
and transitions between these locations (see Fig.~\ref{fig:abstraction-mek-erk}), and a system of timed automata can be
used to model a system of interacting molecules. At any time, each automaton is in a specific location, and together
these locations represent the current state of the biological system. Each timed automaton can have one or more clocks
associated to it, allowing temporal control of transitions between locations. The transitions are used to
represent interactions between molecules. Fast interactions take less time than slow interactions
to perform an activation or inhibition step. We have previously described in detail how approximated
reaction kinetics~\cite{animo-ieee} can be used to calculate the timing of molecular
interactions (see also Suppl. Sect.~\ref{subsec:abstractions}). Figure~\ref{fig:abstraction-mek-erk}
presents a small example that illustrates the basic properties of \tas.
This model describes the activation of ERK by MEK.

>>>>

\subsection*{A brief introduction to ANIMO}
The user interface of ANIMO is displayed in Figure~\ref{fig:cytoscape},
where we present an example of a biological network enriched with dynamic information.
ANIMO is implemented as a plug-in to Cytoscape~\cite{cytoscape}, a software tool developed to represent biological networks.
On top of the static topological information displayed in Cytoscape, ANIMO represents
biological interactions starting from the basic concept of \emph{activity}:
each biological entity in an ANIMO model is considered to be either active or inactive.
Activity is to be interpreted in a very broad sense: for example, an active gene is being transcribed,
an active kinase can perform phosphorylations, etc. Each node in an ANIMO network represents
both active and inactive entities of the same type, with the relative amount of active entities
(the \emph{activity level}) represented by the node colour on a user-configurable scale.
Interactions among nodes define how the biological entities in a network influence each other's activity.
Only nodes whose activity level is larger than 0 (such nodes are called \emph{active}) can have an 
influence on their downstream targets, and only if that influence is not counterbalanced by intervening opposite interactions.
For example, the interaction $A\rightarrow B$ (read ``A activates B'') indicates that node A, if active,
will increase the activity level of node B. If we add an additional interaction to the example,
$C \dashv$\hspace{0.1em}$B$ (``C inhibits B''), with C also active, then the activity level of B will change depending on the
activity levels of A and C, and on their quantitative influence.
The influence of an ANIMO interaction is quantified by a parameter $k$, which defines the speed at which
that interaction occurs: higher values of $k$ give faster interactions. These $k$-values are
the only parameters needed in an ANIMO model, and can be given in either
a quantitative (as real numbers) or a qualitative way, choosing among self-explanatory descriptions
such as ``very slow'', ``slow'', ``medium'', ``fast'', ``very fast''.

@@

\section*{Results}\label{sec:results}
\subsection*{Modelling oscillation in Drosophila Melanogaster circadian clock}\label{sec:animo-drosophila}
To demonstrate that results obtained with ANIMO are comparable to results from widely used modelling
approaches, we present 
an ANIMO model of the circadian clock in \emph{Drosophila Melanogaster} (Suppl. Fig.~\ref{fig:drosophila-model}).
This ANIMO model is based on~\cite{drosophila-ode-model},
where ordinary differential equations (ODEs) were used.
The cyclic behaviour of the circadian clock is based on the alternating formation and destruction of the
CYC/CLK protein complex.
Concentration levels of this complex are in turn regulated by a series of proteins which are produced as
a consequence of CYC/CLK formation. The CWO protein
is central to the functioning of the network, as it degrades the mRNA for most of the involved proteins.
As such, CWO acts as an inhibitor that counterbalances the effect of CYC/CLK.
The positive influence of the light-regulated cryptochrome CRY on the degradation of TIM is a consequence
of the passage between day and night, allowing
the circadian clock to synchronize to a time zone (see Suppl. Sect.~\ref{suppl:repressilator}).

@@

\begin{figure*}[!htpb]
\begin{center}
% \subfloat[\label{fig:large-model-complete}]{\includegraphics[scale=\largeModelScale]{images/large_network_legendg}}\\
%  \subfloat[\label{fig:large-model-graph2}]{\includegraphics[scale=\scalaGrafici]{images/TGF100_vs_TNF100_hyp1_IL-1a4}}
%  \subfloat[\label{fig:large-model-graph4}]{\includegraphics[scale=\scalaGrafici]{images/TNF5_C225_hyp2_JNK1_MK2}}
%  \includegraphics[width=0.9\textwidth]{images/large_network_legendg_completa}
  \includegraphics[width=0.9\textwidth]{Figures/5}
\end{center}
\caption{\csentence{Signalling network downstream of TNF$\alpha$ and EGF in human colon carcinoma cells: improved model.}
{\bf(A)} %{\bf \protect\subref{fig:large-model-complete}}
The model for the merged TNF$\alpha$ and EGF pathways
after addition of the two hypotheses (highlighted).
Hypothesis~1 assumes IL-1$\alpha$ expression to depend on AP-1 activity, which in turn requires
both c-Jun en c-Fos to be activated by JNK1 and ERK, respectively. Hypothesis~2 assumes RAS as an activator
of MEKK1. Node colours represent the activity levels $15$ minutes
after stimulation of 100 ng/ml TNF$\alpha$ + 100 ng/ml EGF.
{\bf(B)} %{\bf \protect\subref{fig:large-model-graph2}}~
After the addition of the first hypothesis (activation of IL-1$\alpha$ production depending both
on JNK1 and ERK): production of IL-1$\alpha$ after stimulation with 100 ng/ml TNF$\alpha$ (series {\sf IL-1a~(TNFa)})
compared with stimulation with 100 ng/ml TGF$\alpha$ (series {\sf IL-1a~(TGFa)}) (24 hours).
{\bf(C)} %{\bf \protect\subref{fig:large-model-graph4}}~
After the addition of the second hypothesis (activation of MEKK1 downstream of EGFR):
activation of JNK1 and MK2 after
stimulation with 5 ng/ml TNF$\alpha$ + 10 $\mu$g/ml C225 (2 hours).
Suppl. Sect.~\ref{suppl:parameters-tnf-egf} explains how the dosage of 5 ng/ml TNF$\alpha$ was represented in the model.\\
The {\sf \_{}data} suffix identifies experimental data; all other series are computed by ANIMO.}\label{fig:large-model-complete}
\end{figure*}

As a second example, we considered the behaviour of JNK1 and MK2. In the model, both
proteins were located downstream of TNF$\alpha$ but not TGF$\alpha$ or EGF. Hence, the
model did not show an effect of C225, a pharmacological inhibitor of ligand-EGFR
binding, on activation of JNK1 or MK2 after stimulation with TNF$\alpha$ (Fig.~\ref{fig:large-model-all}C). However, experimental
data show that C225 strongly reduces activation of JNK1 and MK2 upon stimulation with
TNF$\alpha$~\cite{pathway-autocrine}.
This fact is indicative of a role for EGFR in activation of JNK1 and MK2. Since both JNK1 and MK2
are located downstream of MEKK1, we hypothesized that activation
of MEKK1 is dependent on
both TNF$\alpha$-signalling and TGF$\alpha$-signalling. In the model we added a new
hypothetical node {\sf Hyp~2} (hypothesis~2) to link EGFR to MEKK1 (Suppl. Fig.~\ref{fig:large-model-hypotheses}).
This addition led to an improved fit of the model to the data upon treatment with TNF$\alpha$ + C225:
activation of both MK2 and JNK1 was strongly suppressed by C225 (Fig.~\ref{fig:large-model-complete}C).
Stimulation with EGF alone did not lead to activation of JNK1 and MK2.
These data support the validity of the modification to the model.
Further support for a link between EGFR and MEKK1 was found in literature. Specifically,
Ras has been reported as a direct activator of
MEKK1~\cite{ras-mekk1}. EGFR is a well-known and potent activator of Ras,
which is why it was already in our network~\cite{kegg}.
Other studies also report activation of JNK1 and phosphorylation of c-Jun downstream of Ras, which is consistent with
an interaction between Ras and MEKK1~\cite{cfos-cjun,ras-jnk1}.
Based on these findings, we adapted
our model by removing the {\sf Hyp~2} node and creating a direct interaction between Ras
and MEKK1 (Fig.~\ref{fig:large-model-complete}A). Experimentally, the role of Ras could be confirmed by using a
pharmacological inhibitor of Ras activity, and measuring the effect of this inhibitor on the activation of JNK1 and MK2.
Together, our model suggests that EGFR activity is required
but not sufficient for activation of JNK1 and MK2 in HT-29 cells.

@@

From the point of view of model precision, we position ANIMO between fuzzy logic and ODEs.
Being less parameter-intensive than ODEs and more precise than logic-based models,
ANIMO models are flexible enough to be useful in many occasions.

@@

\def\modelGraphScale{0.148}%0.215}
\def\legendGraphScale{0.16}
\def\halfGraphScale{0.067}%0.1075}
\begin{figure*}[!h]
%\begin{floatingfigure}[l]{.45\textwidth}
\begin{center}
% \subfloat[\label{subfig:mek-erk}]{\includegraphics[scale=.098]{images/abstraction_ta_mek-erk3}}
%   \quad%\qquad\qquad
% \subfloat[\label{subfig:erk}]{\includegraphics[scale=.098]{images/abstraction_ta_erk}}
%   \quad%\qquad\qquad
% \subfloat[\label{subfig:mek}]{\includegraphics[scale=.098]{images/abstraction_ta_mek}}
%   \includegraphics[width=0.7\textwidth]{images/abstraction_ta_mek-erk3_oneImage}
  \includegraphics[width=0.8\textwidth]{Figures/6}
\end{center}
\caption{\csentence{Abstraction of a biochemical reaction to a \tas\ model.}
{\bf(A)} %{\bf \protect\subref{subfig:mek-erk}}~
Classical depiction of a well-studied intracellular signal transduction reaction: protein
MAPK-ERK kinase (MEK) activates downstream protein extracellular-regulated kinase (ERK).
{\bf(B)} %{\bf \protect\subref{subfig:erk}}~
A \ta\ model of ERK, consisting of two locations (circles), {\sf inactive\_ERK} and {\sf active\_ERK},
and one transition (arrow) between the locations. This transition will take place when it is possible to synchronize with
the corresponding action {\sf activate\_ERK!} in the MEK automaton.
{\bf(C)} %{\bf \protect\subref{subfig:mek}}~
A \ta\ model of active MEK, consisting of one location and one
transition. ${\sf t} < 20$ is called an invariant on the location, allowing residence in this location as long as 
clock time {\sf t} is smaller than $20$ units. ${\sf t} > 18$ is called a guard on the transition, allowing the
transition to take place when clock {\sf t} is greater than $18$ units. Together, the invariant and guard in this
example ensure that the transition must take place in the (continuous) time interval $18 < {\sf t} < 20$. When the
transition takes place, the action {\sf activate\_Erk!} is performed (thus allowing the ERK automaton to reach the {\sf
active\_ERK} location) and the local clock coupled to this automaton is reset, ${\sf t} := 0$.\label{fig:abstraction-mek-erk}}
\end{figure*}

@@

Timed Automata have been shown to be a powerful formalism to model biological
processes~\cite{ta-siebert,bartocci-oscillators,oded-ode-ta-discretization}. A timed automaton consists of locations
and transitions between these locations (see Fig.~\ref{fig:abstraction-mek-erk}), and a system of timed automata can be
used to model a system of interacting molecules. At any time, each automaton is in a specific location, and together
these locations represent the current state of the biological system. Each timed automaton can have one or more clocks
associated to it, allowing temporal control of transitions between locations. The transitions are used to
represent interactions between molecules. Fast interactions take less time than slow interactions
to perform an activation or inhibition step. We have previously described in detail how approximated
reaction kinetics~\cite{animo-ieee} can be used to calculate the timing of molecular
interactions (see also Suppl. Sect.~\ref{subsec:abstractions}). Figure~\ref{fig:abstraction-mek-erk}
presents a small example that illustrates the basic properties of \tas.
This model describes the activation of ERK by MEK.
